---
title: "The ANOVA Procedure"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

The following table contains observations $Y_{11}, Y_{12}, \dots, Y_{18}$ and $Y_{21}, Y_{22}, \dots, Y_{28}$ drawn from two normal distributions $N(\mu_1,\sigma^2)$ and $N(\mu_2, \sigma^2)$, respectively.

| $j$ | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| $y_{1j}$  | -0.8 | 0.4 | 2.7 | -0.0 | 0.9 | -2.3 | 1.7 | -0.5 | 
| $y_{2j}$  | 3.7 | 0.6 | 0.3 | 0.4 | -0.7 | 1.1 | 2.1 | 0.2 | 


# Using the ANOVA Procedure

We can test the null hypothesis $H_0: \mu_1 = \mu_2$ against the alternative hypothesis $H_a: \mu_1 \neq \mu_2$ using the ANOVA procedure. Specifically, we can test $H_0$ against $H_a$ using the test statistic 
$$
  F = \frac{MST}{MSE},
$$
which has $F$ distribution with $\nu_1 = 1$ numerator degrees of freedom and $\nu_2 = n - 2 = 14$ denominator degrees of freedom; we obtain an $\alpha$-level test by using the rejection region $F = f_{\alpha, \nu_1, \nu_2}$.

We calculate the mean sum of squares for treatments using the formula
$$
  MST = \frac{n_1}{2} (\bar{y}_1 - \bar{y}_2 )^2,
$$
where $n_1 = n_2 = 8$, and the following code.
```{r calcMST}
# Define data vectors.
y1 <- c(-0.8 , 0.4 , 2.7 , -0.0 , 0.9 , -2.3 , 1.7 , -0.5)
y2 <- c(3.7 , 0.6 , 0.3 , 0.4 , -0.7 , 1.1 , 2.1 , 0.2 )

# Calculate within-group means
y1bar <- mean(y1)
y2bar <- mean(y2)

# Calculate the MST.
n1 <- length(y1)
MST <- (n1/2)*(y1bar - y2bar)^2
```
This yields $\bar{y}_1 = `r y1bar`$, $\bar{y}_2 = `r y2bar`$, and $MST = `r MST`$.

On the other hand, we calculate the sum of squares for error using the following identity
$$
  MSE = \frac{SSE}{2n_1 - 2}, \;\; \text{where }
  SSE = \sum_{i=1}^2 \sum_{j=1}^{n_i} (y_{ij} - \bar{y}_i)^2,
$$
and the following code.
```{r calcMSE}
# Calculate SS for each group of samples.
SSE1 <- sum((y1 - y1bar)^2)
SSE2 <- sum((y2 - y2bar)^2)

# Calculate SSE.
SSE <- SSE1 + SSE2

# Calculate MSE.
MSE <- SSE/(2*n1 - 2)
```
This gives 
\begin{align*}
  \sum_{j=1}^{n_1} (y_{1j} - \bar{y}_1)^2 &= `r SSE1` \\
  \sum_{j=1}^{n_1} (y_{2j} - \bar{y}_2)^2 &= `r SSE2` \\
  SSE &= `r SSE` \\
  MSE &= `r MSE`
\end{align*}

Putting everything together, we calculate the observed value of $F$.
```{r calcF}
fval <- MST/MSE
```
This gives $f = `r fval`$.

Testing with $\alpha = 0.05$, we use the rejection region
$$
  f > f_{\alpha} = `r qf(p = 0.05, df1 = 1, df2 = 14, lower.tail = FALSE)`.
$$
Since the observed value of $f$ is not in the rejection region, we cannot reject $H_0$.

We can calculate the attained level of significance using the following code.
```{r calc-p}
pval <- pf(q = fval, df1 = 1, df2 = 14, lower.tail = FALSE)
```
This gives $p$-value $p = `r pval`$. Since the attained significance is greater than $\alpha = 0.05$, we do not have sufficient evidence to reject $H_0$.

# Using the ``aov`` function

We can also perform this test using the ``aov`` function. To do so, we group our observations as a single vector and create a label vector using the following code.
```{r data-for-aov}
y <- c(y1, y2)
pop <- as.factor(cbind(rep("1", times = 8), rep("2", times = 8)))
```
We next use the command ``aov(formula = y ~ pop)`` to perform the ANOVA procedure.
```{r aov}
aov.mod <- aov(formula = y ~ pop)
```
We can inspect the results of the ANOVA procedure using the ``summary`` function.
```{r aov-summary}
summary(aov.mod)
```
Note that the observed $F$ value and attained significance given by this summary table agree with those calculated earlier.

# Using the ``t.test`` function

Recall that the one-way $F$-test is equivalent to performing a $t$-test for $H_0$ and $H_a$.
We can perform the test using the following code.
```{r t-test}
t.test(y1, y2)
```
It is important to note that the attained significance, $p = 0.3539$, and conclusion is identical to that obtained using the one-sided $F$-test.

