---
title: "ANOVA for Randomized Block Designs in R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



Consider the following data corresponding to a randomized block design with $k = 4$ treatments and $b = 5$ blocks.

```{r data, echo = FALSE}
y <- matrix(data = c(-3.0, -2.3, -0.3, 6.2,
                     -5.0, -2.3, 1.6, 4.6, 
                     -5.5, -1.8, 1.5, 5.5,
                     -6.4, -2.8, -1.6, 5.5,
                     -5.5, -1.6, 0.3, 3.7),
            nrow = 4, ncol = 5)
rownames(y) <- c("Treatment 1", "Treatment 2", "Treatment 3", "Treatment 4")
colnames(y) <- c("Block 1", "Block 2", "Block 3", "Block 4", "Block 5")
knitr::kable(y, align=c(rep('c',times=5)))
```

In order to create the ANOVA table for this experiment, we first calculate the observed values of the sums of squares using the following code.

We start by creating a matrix to store the observations and then calculate CM and Total SS.
```{r SS}
# Create matrix storing observations.
b <- 5
k <- 4
n <- b*k
y <- matrix(data = c(-3.0, -2.3, -0.3, 6.2,
                     -5.0, -2.3, 1.6, 4.6, 
                     -5.5, -1.8, 1.5, 5.5,
                     -6.4, -2.8, -1.6, 5.5,
                     -5.5, -1.6, 0.3, 3.7),
            nrow = 4, ncol = 5)
# Calculate CM and total sum of squares.
cm <- sum(y)^2/n
tss <- sum(y^2) - cm
```

This gives 
\begin{align*}
  \text{CM} &= \frac{(\text{total})^2}{n} = \frac{`r sum(y)^2`}{`r n`} = `r cm` \\
  \text{Total SS} &= \sum_{i=1}^4 \sum_{j=1}^5 (y_{ij} - \bar{y})^2
    = \sum_{i=1}^4 \sum_{j=1}^5 y_{ij}^2 - \text{CM} =
    `r sum(y^2)` - `r cm` = `r tss`.
\end{align*}

Next, we can calculate SSB and SST by noting that $y_{\bullet j}$ and $y_{i \bullet}$ are equal to the $j$th column sum and $i$th row sum of the matrix of observations $y$; the functions ``colSums`` and ``rowSums`` are be used to calculate vectors of column and row sums, respectively for given matrix. We use these functions and the following code to calculate SSB and SST.
```{r SSB-SST}
# Calculate vectors of row and col sums.
ydotj <- colSums(y)
yidot <- rowSums(y)

# Calculate SSB, SST, SSE.
ssb <- sum(ydotj^2)/k - cm
sst <- sum(yidot^2)/b - cm
sse <- tss - ssb - sst
```
This yields 
\begin{align*}
  \text{SSB} &= \sum_{j=1}^5 \frac{y_{\bullet j}^2}{k} - \text{CM} = 
    `r sum(ydotj^2)/k` - `r cm` = `r ssb` \\
  \text{SST} &= \sum_{i=1}^4 \frac{y_{i \bullet}^2}{b} - \text{CM} = 
    `r sum(yidot^2)/b` - `r cm` = `r sst` \\
  \text{SSE} & = \text{Total SS} - \text{SSB} - \text{SST} 
    = `r tss` - `r ssb` - `r sst` = `r sse`.
\end{align*}

# Testing for differences in response to treatments.

Suppose that we want to test a hypothesis of no difference in the mean response for treatments.
We do so using the $F$-distributed statistic
$$
  F = \frac{\text{MST}}{\text{MSE}}
$$
which can be calculated using the following code.
```{r calc-F}
mst <- sst/(k-1)
mse <- sse/(n - b - k + 1)
fval <- mst/mse
```
This gives 
\[
  F = \frac{`r mst`}{`r mse`} = `r fval`.
\]
For a $\alpha$-level test we use the critical value of the $F$-statistic for $\nu_1 = k-1 = `r k - 1`$ and $\nu_2 = n-b-k+1  =`r n-b-k+1`$ df.
For $\alpha = 0.05$ we have 
$$
  f_{0.05} = `r round(qf(p = 0.05, df1 = k-1, df2 = n-b-k+1, lower.tail = FALSE), digits = 4)`.
$$
Since the observed value of $F$ is in the rejection region, we can infer that there is a significant difference between mean response for some of the treatments.

# Testing for differences between blocks.

Similarly, we can test the hypothesis $H_0: \beta_i = 0$ for all $i = 1,2, \dots, 5$ using the test statistic 
$$
  F = \frac{\text{MSB}}{\text{MSE}},
$$
which we calculate using the following code.
```{r f-blocks}
msb <- ssb/(b-1)
fval <- msb/mse
```
This gives 
$$
  F = \frac{`r msb`}{`r mse`} = `r fval`.
$$
The statistic $F$ has $f$-distribution with $b-1 = `r b -1`$ numerator degrees of freedom, $n - b- k +1 =`r n-b-k+1`$ denominator degrees of freedom. The critical value for an $\alpha$-level test, with $\alpha = 0.05$, is given by 
$$
  f_{0.05} = `r round(qf(p = 0.05, df1 = b-1, df2 = n-b-k+1, lower.tail = FALSE), digits = 4)`.
$$
Therefore, we cannot reject $H_0$ since the observed value of $F$ is outside the rejection region.


# Using the ``aov`` function

We can also use the ``aov`` function to calculate the ANOVA table and test whether there is a significant difference in mean response to treatments or across blocks.

We first create vectors of labels indicating which treatment and block each observation corresponds to. We do so using the ``gl`` function in R.
```{r create-treatment-labels}
f <- c("Treatment 1", "Treatment 2", "Treatment 3", "Treatment 4")
ts <- gl(n = k, k = 1, length = n, labels = factor(f))
```
The argument ``n = k`` indicates to create a vector with ``k`` labels (contained in the vector ``f``), while the argument ``k = 1`` indicates to list each label once in sequence.
This yields a $k$-dimensional factor vector with the sequence *Treatment 1*, *Treatment 2*, *Treatment 3*, *Treatment 4* repeated $b = 5$ times indicating the treatment applied to each observation $y_{ij}$ when the matrix $y$ is treated as a $n$-dimensional vector.

Similarly, the following code generates the label vector assigning blocks to each observation.
```{r create-block-labels}
bs <- gl(n = b, k = k, length = n)
```
The argument ``k = k`` indicates to repeat each factor level (here, $1,2,\dots, b$ by default) a total of $k$ times while creating the vector ``bs``.
This yields vector ``bs`` consisting of $k = 4$ copies of $1$, followed by $k$ copies of $2$, etc. The first $k$ entries correspond to the observations in the first column of $y$, which all belong to block $1$, the second $k$ correspond to the second column of $y$ (block $2$), and so we have a vector associating blocks with each observation.

After creating these label vectors we use the ``aov`` function with model ``y ~ ts + bs`` to create the desired ANOVA table.
```{r anova-tab}
summary(aov(as.vector(y) ~ ts + bs))
```


