---
title: "ANOVA for One-Way Layouts in R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The following data are drawn from four normal populations.
```{r data, echo=FALSE}
y1 <- c(0.9,  0.4,  0.7,  0.2,  0.3,  0.0,  0.6, -1.4)
y2 <- c(-1.1,  0.7, -0.5,  2.0,  2.3, -0.7,  1.6,  1.0)
y3 <- c(2.3,  0.5, -0.2,  0.5,  1.9,  1.9,  2.4,  2.0)
y4 <- c( -1.0, -0.3, -0.7,  0.6,  3.3,  3.0,  2.5,  0.0)

dta <- data.frame(y1, y2, y3, y4)

knitr::kable(dta, 
             col.names = c("Pop 1", "Pop 2", "Pop 3", "Pop 4"),
             align=c(rep('c',times=4)))
```
We can store the given data as a data frame using the following code. Note that we can do so because each sample has the same size ($n_i = 8$).
```{r store-as-frame}
y1 <- c(0.9,  0.4,  0.7,  0.2,  0.3,  0.0,  0.6, -1.4)
y2 <- c(-1.1,  0.7, -0.5,  2.0,  2.3, -0.7,  1.6,  1.0)
y3 <- c(2.3,  0.5, -0.2,  0.5,  1.9,  1.9,  2.4,  2.0)
y4 <- c( -1.0, -0.3, -0.7,  0.6,  3.3,  3.0,  2.5,  0.0)

dta <- data.frame(y1, y2, y3, y4)
colnames(dta) <- c("1", "2", "3", "4")
```

The following code calculates the total and averages for each population.
```{r tots-means}
Ysums <- colSums(dta)
Ymeans <- colMeans(dta)
```

This gives 
\begin{align*}
  y_{1\bullet} &= `r Ysums[1]`, &&& y_{2\bullet} &= `r Ysums[2]`,
  &&& y_{3\bullet} &= `r Ysums[3]`, &&& y_{4\bullet} &= `r Ysums[4]`, \\
  \bar y_{1\bullet} &= `r Ymeans[1]`, &&& \bar y_{2\bullet} &= `r Ymeans[2]`,
  &&& \bar y_{3\bullet} &= `r Ymeans[3]`, &&& \bar y_{4\bullet} &= `r Ymeans[4]`
\end{align*}


To test the null hypothesis $H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4$, we need to calculate the MST and MSE for this data. We start by calculating the correction for the mean CM using the following code.
```{r cm}
n <- 32 # Total number of samples.
cmval <- sum(dta)^2/n
```
This gives 
\[
  \text{CM} = \frac{1}{n}\left( \sum_{i=1}^k \sum_{j=1}^{n_i} y_{ij} \right)^2
    = \frac{(`r sum(dta)`)^2}{`r n`}
    = `r cmval`.
\]

We can calculate SST and MST using the formulas
\[
  \text{SST} = \sum_{i=1}^k \frac{y_{i\bullet}^2}{n_i} - \text{CM},
  \hspace{0.25in} \text{MST} = \frac{\text{SST}}{k-1},
\]
and the following code.
```{r mst}
k <- 4 # number of populations.
sst <- sum(Ysums^2/8) - cmval
mst <- sst/(k-1)
```
This gives 
$$
  \text{SST} = `r sst`, \hspace{0.5in}
  \text{MST} = `r mst`.
$$




On the other hand, we calculate MSE using the following formulas
$$
  \text{TSS} = \sum_{i=1}^k \sum_{j=1}^{n_i} y_{ij}^2 - CM,
  \hspace{0.25in}
  \text{SSE} = \text{TSS} - \text{SST},
  \hspace{0.25in}
  \text{MSE} = \frac{\text{SSE}}{n-k}
$$
and the following code.
```{r mse}
tss <- sum(dta^2) - cmval
sse <- tss - sst
mse <- sse/(n - 4)
```
This yields
$$
 \text{TSS} = `r tss`,
  \hspace{0.25in}
  \text{SSE} = `r sse`,
  \hspace{0.25in}
  \text{MSE} = `r mse`.
$$


Finally, we calculate the observed value of the test statistic $F = \text{MST}/\text{MSE}$ using the following code.
```{r fval}
fval <- mst/mse
```
This gives $f = `r fval`$, with $\nu_1 = 3$ and $\nu_2 = 28$ degrees of freedom.
The rejection region for a $\alpha = 0.05$ level test is
$$
  f > f_{\alpha} = 
  `r qf(p = 0.05, df1 = 3, df2 = 28, lower.tail = FALSE)`.
$$
Since the observed value of $F$ is not in the rejection region, we cannot reject $H_0$.

 
# Using the aov function.

We can also calculate the ANOVA table using the ``aov`` function in R.
To do, we first must store our data as a two-dimensional array with one column indicating population member and the second indicating the observed response (as opposed to four columns giving the responses for each sample).

We can do so using the ``melt`` function from the ``reshape2`` package in R. After installing ``reshape2`` using the command ``install.packages(reshape2)``, we can perform the necessary transformation of the data using the following code.
```{r meltdata}
library(reshape2)
meltedY <- melt(dta)
colnames(meltedY) <- c("Population", "Response")
```
```{r displayMeltedY, echo = FALSE}
knitr::kable(meltedY,align=c(rep('c',times=2)))
```
Finally, we can form the ANOVA table for this data using the ``aov`` and ``summary`` functions in R.
```{r formANOVATable}
resAOV <- aov(Response ~ Population, data = meltedY)
summary(resAOV)
```

Note that the values found in the ANOVA table given by the ``aov`` agree with those calculated above.

# Calculating Confidence Intervals

We can make inferences about the individual populations using their sample means and the pooled estimator of $\sigma^2$ given by the MSE.

For example, suppose we want to test the hypothesis $H_0: \mu_3 = 0$ against the alternative hypothesis $H_a: \mu_3 \neq 0$.
We can calculate a two-sided 95% confidence interval for $\mu_3$ from this data using the formula
$$
  \bar{Y}_{3\bullet} \pm t_{0.025} \sqrt{ \frac{-\text{MSE}}{n_3}},
$$
where $t_{0.025}$ is calculated using $n-k$ degrees of freedom.

We calculate the confidence interval using the following code.
```{r CI}
tval <- qt(p = 0.025, df = n - k, lower.tail = FALSE)
s <- sqrt(mse)
rad <- tval*s/sqrt(8)
lb <- Ymeans[3] - rad
ub <- Ymeans[3] + rad
```
This gives the confidence interval
$$
  `r lb ` < \mu_3 < `r ub`.
$$
Since this confidence interval *does not contain 0* we can reject $H_0$.

As a second example, suppose that we want to test the null hypothesis $H_0: \mu_2 = \mu_4$ against $H_a: \mu_2 \neq \mu_4$.
We can calculate a 95\% confidence interval for the difference $\mu_2 - \mu_4$ using the formula
$$
  \bar{Y}_{2\bullet} - \bar{Y}_{4\bullet} \pm t_{0.025} \sqrt{\text{MSE}}\sqrt{\frac{1}{n_2} + \frac{1}{n_4}}
$$
and the following code.
```{r diffCI}
s <- sqrt(mse)
rad <- tval*s*sqrt(1/8 + 1/8)
lb <- Ymeans[2] - Ymeans[4] - rad
ub <- Ymeans[2] - Ymeans[4] + rad
```
This gives the confidence interval
$$
  `r lb ` < \mu_2 - \mu_4 < `r ub`.
$$
This implies that we cannot reject $H_0$ based on this data since $0$ belongs to this confidence interval.