---
title: "Analyzing Categorical Data using the Pearson Statistic and R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE)
```

Consider a multinomial experiment with $k = 4$ possible outcomes.
After $n = 100$ trials are performed, we observe the following counts of each outcome.

```{r exData, echo = FALSE}
y <- c(20, 17, 31, 32)
n <- sum(y)

dta <- data.frame(1:4, y)
colnames(dta) <- c("Outcome", "Observed Count")

knitr::kable(dta, row.names = FALSE, align = "cc", booktabs = TRUE) %>% 
  kable_styling(full_width = FALSE) 
```

Suppose that we wanted to determine whether the probabilities of each outcome differ; this analysis will echo that of Example 14.1.
More precisely, we want to test the null hypothesis $H_0: p_1 = p_2 = p_3 = p_4 = 0.25$.
To do so, we use the Pearson test statistic $X^2$ with approximate $\chi^2$ distribution; in this case, $X^2$ has $k-1$ degrees of freedom since the only assumption used about the outcome probabilities is that $p_1 + p_2 + p_3 + p_4 = 1$.

We can calculate $X^2$ using the following code.
```{r calc-Test-Stat}
# Set outcome probabilities under H0.
p <- rep(0.25, 4) 

# Define vector of observed counts.
y <- c(20, 17, 31, 32)
n <- sum(y)

# Calculate X^2.
Xsq = sum((y - p*n)^2/(p*n))
```

This gives $X^2 = `r Xsq`$.

We have an $\alpha$-level test if we reject $H_0$ when $X^2 > \chi^2_{\alpha, k-1}$.
For $\alpha = 0.05$ and $k-1 = 3$, we have the rejection region
$$
  X^2 > `r round(qchisq(p = 0.05, df = 3, lower.tail = FALSE), digits = 3)`.
$$
In this case, we cannot reject $H_0$ since the observed value of $X^2$ is outside the rejection region.

# Using the ``chisq.test`` function

We could have also performed the test using the ``chisq.test`` function, as in the following code.

```{r}
chisq.test(y)
```

When called, ``chisq.test(y)`` calculates the value of $X^2$ for the observed counts stored in the vector ``y``, as well as the observed attained significance level. Here, we have $p = 0.07318$ which indicates that we cannot reject $H_0$ (using $\alpha = 0.05$).