---
title: "Confidence Intervals in Randomized Block Designs using R"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

Consider the following data corresponding to a randomized block design with $k = 4$ treatments and $b = 5$ blocks.

```{r data, echo = FALSE}
y <- matrix(data = c(-3.0, -2.3, -0.3, 6.2,
                     -5.0, -2.3, 1.6, 4.6, 
                     -5.5, -1.8, 1.5, 5.5,
                     -6.4, -2.8, -1.6, 5.5,
                     -5.5, -1.6, 0.3, 3.7),
            nrow = 4, ncol = 5)
rownames(y) <- c("Treatment 1", "Treatment 2", "Treatment 3", "Treatment 4")
colnames(y) <- c("Block 1", "Block 2", "Block 3", "Block 4", "Block 5")
knitr::kable(y, align=c(rep('c',times=5)))
```

We calculated the following the sums of squares in an earlier example:
```{r SS, echo = FALSE}
# Create matrix storing observations.
b <- 5
k <- 4
n <- b*k
y <- matrix(data = c(-3.0, -2.3, -0.3, 6.2,
                     -5.0, -2.3, 1.6, 4.6, 
                     -5.5, -1.8, 1.5, 5.5,
                     -6.4, -2.8, -1.6, 5.5,
                     -5.5, -1.6, 0.3, 3.7),
            nrow = 4, ncol = 5)
# Calculate CM and total sum of squares.
cm <- sum(y)^2/n
tss <- sum(y^2) - cm

# Calculate vectors of row and col sums.
ydotj <- colSums(y)
yidot <- rowSums(y)

# Calculate SSB, SST, SSE.
ssb <- sum(ydotj^2)/k - cm
sst <- sum(yidot^2)/b - cm
sse <- tss - ssb - sst

# MSE/MST/MSB
msb <- ssb/(b-1)
mst <- sst/(k-1)
mse <- sse/(n - b - k + 1)
```

\begin{align*}
  \text{CM} &= `r cm` &&&
  \text{Total SS} &= `r tss` \\ 
  \text{SSB} &= `r ssb` &&& \text{MSB} &= `r msb` \\
  \text{SST} &= `r sst` &&& \text{MST} &= `r mst` \\
  \text{SSE} &= `r sse` &&& \text{MSE} &= `r mse`
\end{align*}

# Comparing treatments

Recall that a $100(1-\alpha)\%$-confidence for the difference between $\tau_1$ and $\tau_2$ is given by 
$$
  \bar{Y}_{1\bullet} - \bar{Y}_{2\bullet} \pm t_{\alpha/2} S \sqrt{\frac{2}{b}},
$$
where $t_{\alpha/2}$ is calculated using $n - b - k + 1 = 12$ degrees of freedom and $S = \sqrt{MSE}$ gives an unbiased estimate of $\sigma^2$.

We can calculate the observed estimate of $\tau_1 - \tau_2$ by using the ``rowMeans`` function to calculate $\bar{Y}_{i\bullet}$ for each $i = 1,2,3,4$.
```{r rowMeans}
# Create matrix storing observations.
b <- 5
k <- 4
n <- b*k
y <- matrix(data = c(-3.0, -2.3, -0.3, 6.2,
                     -5.0, -2.3, 1.6, 4.6, 
                     -5.5, -1.8, 1.5, 5.5,
                     -6.4, -2.8, -1.6, 5.5,
                     -5.5, -1.6, 0.3, 3.7),
            nrow = 4, ncol = 5)

# Calculate row averages.
rm <- rowMeans(y)
```
The first two entries of the output vector ``rm`` store $\bar{Y}_{1\bullet}$ and $\bar{Y}_{2\bullet}$ respectively. The following code yields the desired confidence interval for $\alpha = 0.05$.
```{r tau-ci}
alpha <- 0.10
rad <- qt(p = alpha/2, df = n - b - k + 1, lower.tail = FALSE)*sqrt(mse)*sqrt(2/b)
lb <- rm[1] - rm[2] - rad
ub <- rm[1] - rm[2] + rad
```

This gives the $90\%$ confidence interval for $\tau$
$$
  `r rm[1]` + `r - rm[2]` \pm 
  `r round(qt(p = alpha/2, df = n - b - k + 1, lower.tail = FALSE), digits = 4)` (`r sqrt(mse)`) \sqrt{\frac{2}{5}}
$$
or, equivalently
$$
  `r lb` < \tau_1 - \tau_2 < `r ub`.
$$

Note that we can conclude that there is a significant evidence that $\tau_1$ and $\tau_2$ differ (at the $\alpha = 0.05$ level) since $0$ is outside this confidence interval.