---
title: "Testing for Correlation"
output:
  html_notebook: default
  pdf_document: default
---

The following table provides the weight $W$ (in pounds) and length $l$ (in inches) for $15$ alligators in central Florida. Because length is easier observe than weight, we want to construct a model relating weight to length, which could be used to predict the weights of alligators of specified lengths. In particular, we want to fit the model 
$$
  \ln W = \ln \alpha_0 + \alpha_1 \ln l + \epsilon = \beta_0 + \beta_1 x + \epsilon.
$$
Let's test if there is a correlation between $X =\ln l$ and $Y= \ln W$.

| Alligator | $x = \ln l$| $y = \ln W$ |
|:---:|:---:|:---:|
|1| 3.87| 4.87 |
|2| 3.61 |3.93 |
|3 |4.33 |6.46 |
|4 |3.43 |3.33 |
|5 |3.81 |4.38 |
|6 |3.83 |4.70 |
|7 |3.46 |3.50 |
|8| 3.76 |4.50 |
|9| 3.50 |3.58 |
|10| 3.58 |3.64|
|11| 4.19 |5.90|
|12| 3.78 |4.43|
|13| 3.71 |4.38|
|14| 3.73 |4.42|
|15| 3.78 |4.25|

# Calculating the Test Statistic

We first create vectors to store the observed values of $X$ and $Y$.
```{r create_data}
x = c(3.87, 3.61, 4.33, 3.43, 3.81, 3.83, 3.46, 3.76, 3.5, 3.58, 4.19,
      3.78, 3.71, 3.73, 3.78)
y = c(4.87, 3.93, 6.46, 3.33, 4.38, 4.70, 3.5, 4.5, 3.58, 3.64, 5.9, 4.43, 4.38, 4.42, 4.25)
```

We can calculate $S_{xx}, S_{xy},$ and $S_{yy}$, which will be used to calculate sample correlation coefficient $r$, using the following code.
```{r calc_summary_stats}
# Number of observations.
n <- length(x)

# Sums.
sumx <- sum(x = x)
sumy <- sum(x = y)
sumxy <- sum(x = x*y)
sumxx <- sum(x = x^2)
sumyy <- sum(x = y^2)

# Calculate Sxy and Sxx
Sxy <- sumxy - 1/n*sumx*sumy
Sxx <- sumxx - 1/n*sumx^2
Syy <- sumyy - 1/n*sumy^2
```

This yields 
$$
\begin{array}{lll}
  \sum x_i = `r sumx` & \sum x_i^2 = `r sumxx` & S_{xx} = `r Sxx` \\
  \sum y_i = `r sumy` & \sum y_i^2 = `r sumyy` & S_{yy} = `r Syy` \\
    & \sum x_i y_i = `r sumxy` & S_{xy} = `r Sxy`
\end{array}
$$
We can calculate the sample correlation coefficient using the formula 
$$
  r = \frac{S_{xy}}{\sqrt{S_{xx} S_{yy}}}
$$
and the following code.
```{r calcR}
r <- Sxy/sqrt(Sxx*Syy)
```
This gives $r = `r r`$.

# Performing the test

The sample correlation coefficient $r$ has value close to $1$, which suggests a strong positive correlation between the logarithms of length and width. 
Assuming that $(X,Y)$ has a bivariate normal distribution, we can test whether $X$ and $Y$ are independent using the hypotheses $H_0: \rho = 0$ and $H_a: \rho \neq 0$.
We can calculate the value of the test statistic using the formula
$$
  t = \frac{r \sqrt{n-2}}{\sqrt{1 - r^2}}
$$
and the following code.
```{r calcT}
tval <- r*sqrt(n-2)/sqrt(1-r^2)
```
This gives $t = `r tval`$. 
Testing at $\alpha = 0.05$ level of significance, we have rejection region given by $|t| > t_{0.025} = `r qt(0.025, df = n-2, lower.tail = FALSE)`$. Since the observed value of $t$ lies in the rejection region, we can reject $H_0$ and conclude that we have evidence that $X$ and $Y$ are not independent.
The following code calculates the attained significance level.
```{r pval}
2*pt(q = tval, df = n - 2, lower.tail = FALSE)
```

# Coefficient of Determination

The coefficient of determination $r^2$ is given by following code.
```{r CoD}
r^2
```



