---
title: "Inferences involving the Linear Functions of the Parameters"
output:
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
```


We previously considered the following observations $Y_1, Y_2, \dots, Y_n$ following linear model $E(Y) = \beta_0 + \beta_1 x$: 

| $x$ | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
|$y$ | 4.13 |  4.78 | 6.44 |  5.45 | 7.99 |  7.99 |  8.95 | 11.02 | 11.89 | 10.75 |

In the previous example, we fit the least squares model 
$$
  \hat y = 3.18 + 0.865 x
$$
The following code plots each point $(x,y)$ in the plane as well as the least squares model line.
We can plot the fitted line, as well as the observed data using the **abline** function, which plots a line with given slope and intercept.
```{r plot fitted line}
x <- 1:10
y <-  c(4.13,  4.78,  6.44,  5.45,  7.99,  7.99,  8.95, 11.02, 11.89, 10.75)
plot(x, y, xlim = c(0,10), ylim = c(0,12), col = "navy", pch = 19, cex = 1.5)
abline(3.18, 0.865, col = "red", lwd = 2)
```

# A Confidence Interval for Expected Value

We want to calculate a $100(1-\alpha)\%$ confidence interval for $E(Y) = \beta_0 + \beta_1 x^*$.
This is a special case of $a_0 \beta_0 + a_1 \beta_1$ with $a_0 = 1$ and $a_1 = x^*$. Thus, we can use the formula for the general confidence interval, specialized for this particular choice of $a_0, a_1$.

As an example, let's calculate a $90\%$ confidence interval for $E(Y)$ at $x^* = a_1 = 1$ for this sample.
Using the formula derived earlier, the desired confidence interval is given by the formula
$$
  \hat \beta_0 + \hat \beta_1 \pm t_{0.05} S 
  \sqrt{\frac{1}{n} + \frac{(x^* - \bar x)^2}{S_{xx}}},
$$
where $t_{0.05}$ is based on $n-2 = 8$ degrees of freedom.

We calculated many of the quantities in the formula earlier, which we load using the following code.
```{r}
# Model parameters.
hb0 <- 3.18
hb1 <- 0.865

# Recalculate mean, Sxx.
n <- length(x)
xbar <- mean(x)
Sxx <- sum(x^2) - n*xbar^2

# Estimate of the variance.
s <- sqrt(0.6515)
```


Finally, we obtain the confidence by substituting $x^* = 1$ using the following code.
```{r Calculate interval}
xstar <- 1 

# Calculate the radius/half-width of the interval.
rad <- qt(p = 0.05, df = n - 2, lower.tail = FALSE)*s*sqrt(1/n + (xstar - xbar)^2/Sxx)

# Calculate confidence interval.

lb <- hb0 + hb1*xstar - rad
ub <- hb0 + hb1*xstar + rad
```

This yields the confidence interval 
$$
  `r lb` < E(Y) < `r ub`.
$$

We can superimpose this confidence interval on our earlier plot of the observations and least squares line using the following code.
```{r}
# Earlier plot.
plot(x, y, xlim = c(0,10), ylim = c(0,12), col = "navy", pch = 19, cex = 1.5)
abline(3.18, 0.865, col = "red", lwd = 2)
# Superimpose confidence interval.
lines(c(1,1), c(lb, ub), col = "black", lwd = 3)
```
Note that the observed value of $Y$ corresponding to $x^* = 1$ lies within this confidence interval.

# Prediction Intervals

Suppose that we want to predict the value of $Y$ for a given value of $x = x^*$. We know that a $100(1-\alpha)\%$ \emph{prediction} interval for $Y$ is given 
$$
  \hat\beta_0 + \hat\beta_1 x^* \pm t_{\alpha/2} S \sqrt{1 + \frac{1}{n} + \frac{(x^* - \bar x)^2}{S_{xx}}}.
$$
Suppose that we want to predict the value of $Y$ for $x = 1$. For example, this would occur if we run the underlying experiment again using the independent variable value $x = 1$ (we've already generated one observation $Y = 4.13$ corresponding to $x = 1$). 
We can generate a $90\%$ prediction interval for $Y$ using the following code.

```{r predInt}
# Choose x and alpha.
xstar <- 1
alpha <- 0.1

# Calculate radius of the prediction interval.
prad <- qt(p = alpha/2, df = n -2, lower.tail = FALSE)*s*sqrt(1 + 1/n + (xstar - xbar)^2/Sxx)

# Calculate endpoints of the prediction interval.
plb <- hb0 + hb1*xstar - prad
pub <- hb0 + hb1*xstar + prad
```
This gives the prediction interval 
$$
  `r plb` < Y < `r pub`,
$$
with width $`r 2*prad`$. Note that the code for generating the prediction interval is nearly identical to that used to calculate the confidence interval for $E(Y)$ at $x = 1$; the only difference is the presence of the $1$-term in the square root of the formula for the prediction interval. We should note further that the prediction interval is significantly wider than the confidence interval for $E(Y)$.


# Plotting Confidence and Prediction Bands

The formulas for confidence interval for $E(Y)$ at $x^*$ and prediction interval for $Y$ at $x^*$ can be evaluated for a \emph{set} of $x^*$ with minor modification to our earlier code.
For example, the following code generates the prediction intervals for $Y$ for each $x^*$ in the set $\{0, 0.25, 0.5, \dots, 11.75, 12\}$.

```{r predBandsSet}
# Choose x and alpha.
xstar <- seq(from = 0, to = 12, by = 0.25)
alpha <- 0.1

# Calculate radius of the prediction interval.
prads <- qt(p = alpha/2, df = n -2, lower.tail = FALSE)*s*sqrt(1 + 1/n + (xstar - xbar)^2/Sxx)

# Calculate endpoints of the prediction interval.
plbs <- hb0 + hb1*xstar - prads
pubs <- hb0 + hb1*xstar + prads
```

The only difference between this code and that used to calculate the prediction at a single point earlier, is that we now evaluate the formula for the endpoint with $x^*$ given as an input vector (rather than a single value). The output of this code is vectors **plbs** and **pubs** storing the lower and upper bounds, respectively, of the prediction interval for each value of $x^*$  upper bound.

Similarly, we can generate confidence intervals for $E(Y)$ at each $x^*$ in the same set using the following code.

```{r CIs}
rads <- qt(p = alpha, df = n - 2, lower.tail = FALSE)*s*sqrt(1/n + (xstar - xbar)^2/Sxx)
lbs <- hb0 + hb1*xstar - rads
ubs <- hb0 + hb1*xstar + rads
```

We can use these calculated intervals to plot confidence and prediction bands for the least-squares line.
```{r bands}
# Earlier plot.
plot(x, y, xlim = c(0,10), ylim = c(0,12), col = "navy", pch = 19, cex = 1.5)
abline(3.18, 0.865, col = "red", lwd = 2)

# Plot curve indicating lower/upper bounds on confidence intervals.
lines(xstar, lbs, col = "black", lty = 2, lwd = 2)
lines(xstar, ubs, col = "black", pch = "o", lty = 2, lwd = 2)

# Plot prediction band.
lines(xstar, plbs, col = "blue", pch = "o", lty = 3, lwd = 2)
lines(xstar, pubs, col = "blue", pch = "o", lty = 3, lwd = 2)
```

